### 2016-10-04 Paper_v8.tex
PLoS formatting

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{lineno}

\usepackage{listings}

\urlstyle{rm}

%\textwidth = 6.50 in
%\textheight = 9.5 in
%\oddsidemargin =  0.0 in
%\evensidemargin = 0.0 in
%\topmargin = -0.50 in
%\headheight = 0.0 in
%\headsep = 0.25 in
%\parskip = 0.15in
%\linespread{1.75}
\doublespace

\setallmargins{1in}

%\bibliographystyle{plain}
\bibliographystyle{msb}

\definecolor{mygreen}{rgb}{0,0.6,0}
\definecolor{mygray}{rgb}{0.5,0.5,0.5}
\definecolor{mymauve}{rgb}{0.58,0,0.82}

\lstdefinestyle{myCustomMatlabStyle}{
%belowcaptionskip=\baselineskip,
breaklines=true,
%frame=L,
xleftmargin=\parindent,
language=Octave,
showstringspaces=false,
basicstyle=\tiny\ttfamily,
keywordstyle=\bfseries\color{black},
commentstyle=\itshape\color{black},
identifierstyle=\color{blue},
stringstyle=\color{orange},
}

\makeatletter
\renewcommand\subsection{\@startsection
	{subsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\bfseries}}
\renewcommand\subsubsection{\@startsection
	{subsubsection}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\renewcommand\section{\@startsection
	{subsection}{2}{0mm}
	{-0.2in}
	{0.05\baselineskip}
	{\normalfont\large\bfseries}}
\renewcommand\paragraph{\@startsection
	{paragraph}{2}{0mm}
	{-0.05in}
	{-0.5\baselineskip}
	{\normalfont\normalsize\itshape}}
\makeatother

%Review style settings
%\newenvironment{bmcformat}{\begin{raggedright}\baselineskip20pt\sloppy\setboolean{publ}{false}}{\end{raggedright}\baselineskip20pt\sloppy}

%Publication style settings

% Single space'd bib -
\setlength\bibsep{0pt}

\renewcommand{\rmdefault}{phv}\renewcommand{\sfdefault}{phv}

% Change the number format in the ref list -
\renewcommand{\bibnumfmt}[1]{#1.}

% Change Figure to Fig.
\renewcommand{\figurename}{Fig.}

\begin{document}

\begin{titlepage}
{\par\centering\textbf{\Large Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\vspace{0.05in}
{\par \centering Russell Gould$^{\ddag}$ and David M. Bassen, Anirikh Chakrabarti, Jeffrey D. Varner and Jonathan Butcher$^{\ddag *}$}
\vspace{0.05in}
{\par \centering Robert Frederick Smith School of Chemical and Biomolecular Engineering and $^{\ddag}$Nancy E. and Peter C. Meinig School of Biomedical Engineering, Cornell University, Ithaca NY 14853}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~ Modeling of TGF-$\beta$ induced EMT}
\vspace{0.1in}
% {\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology (60,000 Character Limit)}}
%\vspace{0.1in}
{\par \centering \textbf{Character count:}~\emph{56,659}}
\vspace{0.1in}
{\par \centering $^{*}$Corresponding author:}
\vspace{0.1in}
{\par \centering Jonathan T. Butcher,}
{\par \centering Associate Professor, Nancy E. and Peter C. Meinig School of Biomedical Engineering,}
{\par \centering 304 Weill Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jtb47@cornell.edu}
{\par \centering Phone: (607) 255 - 3575}
{\par \centering Fax: (607) 255 - 7330}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
We used ordinary differential equations (ODEs) to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was enhanced following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

%as seen by the initiating the formation of endocardial cushions in the atrioventricular valves and mediating palate fusion of medial epithelial cells \citep{Zavadil:2005fu}.  In this context, EMT terminates and has been suggested as a highly controlled process \citep{Lee:2006tw}.  EMT is also important under pathological conditions, prominently in fibrosis and in invasion and metastasis of carcinomas \citep{Thiery:2003il}. This culminates decades of work within the EMT field appears shared in developmental and pathological conditions \citep{Xu:2009oq}.
%In this study, we determined the extent of EMT in normal epithelial cells exposed to pathological conditions by (1) implementing and subsequently interrogating a population based model containing canonical EMT regulatory networks, and (2) verify our mechanistic targets using established cell lines with differing induction capacities for EMT. Toward this objective, we modeled the core molecular interactions of EMT using mass-action kinetics within an ordinary differential equation (ODE) framework.  A population approach was used to estimate a family or ensemble of EMT models that was consistent with the previous literature \citep{Medici:2006qa,Medici:2008fk}.  Using the newly predicted operational paradigms, we sought to capture the role of phospho-Sp1 and NFAT on phenotype heterogeneity during TGF$\beta$ induced EMT \emph{in vitro} via key marker expression profiles and fate processes.
%Infrastructure components include TGF-$\beta$2 ligand binding, TGF-$\beta$RI,II,III transmembrane growth factor receptor phosphorylation, activation of the Ras/MAPK cascade, cytosolic release of $\beta$-catenin from E-cadherin, autocrine response of the TGF-$\beta$3 ligand, phosphorylation of the Smad proteins, and cross-talk between the MAPK/Smad signaling pathways.
%Furthermore, because these reactions occur with varying initial conditions and unknown parameter values, significant  differences may exist in the overall response of the model.
%To unravel the operation of this complex architecture, we have used mathematical modeling.
%In this program, extracellular TGF-$\beta$1/2 induces a series of  that ultimately repress epithelial gene expression
%while simultaneously activating mesenchymal gene expression.
%However, this program is not a simple switch. It involves
%Thus, the molecular architecture responsible for EMT induction is highly complex, involving multiple inputs and overlapping transcriptional networks.
%However, including these additional mechanisms increases model complexity.
%Analysis of the model predicted that the NFATc and phosphorylated Sp1 transcription factors were likely the key elements driving this heterogeneity.

@@

\section*{Introduction}
The epithelial to mesenchymal transition (EMT) is a broadly participating, evolutionarily conserved differentiation program essential for tissue morphogenesis, remodeling and pathological processes such as cancer \citep{Thiery:2003il}. During EMT polarized, tightly adhered epithelial cell monolayers are transformed into non-interacting motile mesenchymal cells that simultaneously degrade and synthesize extracellular matrix (ECM) components and invade into the underlying tissue space \citep{Stahl:2001pd}. EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \citep{Eisenberg:1995ai} (also Kalluri J Clin Invest 2009, Thiery Cell 2009). Transforming growth factor $\beta$ (TGF-$\beta$) family members are important inducers of both developmental and pathological EMT \citep{Zavadil:2005fu,Xu:2009oq}. Decades of research has focused on identifying molecular regulators of EMT, but almost all on a single gene and in a nearly binary yes/no level of qualitative understanding. Medici and coworkers identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \citep{Medici:2006qa,Medici:2008fk}.
This program involves carefully orchestrated rounds of gene expression driven by the Smad and Snail families of transcription factors as well as
other key factors such as lymphoid enhancer-binding factor 1 (LEF-1), nuclear factor of activated T-cells, cytoplasmic 1 (NFATc1), and specificity protein 1 (Sp1).
Coregulators such as $\beta$-catenin, NF-$\kappa$B, and the ErbB family of receptor tyrosine kinases however also participate in EMT regulation, but the degree of each's influence is difficult to ascertain in isolation \citep{Kim:2002lh,Jiang:2007aa,Huber:2004aa,Hardy:2010aa}.
EMT also exhibits complex temporal dynamics that are often intractable in gain/loss of function studies.
Elucidating the master regulatory architecture controlling EMT therefore requires inclusion of these complex overlapping and non-binary behaviors.

Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \citep{Ahmed:2007mi}.
Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \citep{Chung:2009jl}.
Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \citep{Vilar:2006gb}.
Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
Data-driven systems approaches \citep{Cirit:2012kx} or logical model formulations \citep{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \citep{2001_bailey_NatBiotech}.
Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \citep{Machta:2013by}.
Thus, reasonable model predictions are often possible with only limited parameter information.
Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \citep{Song:2010fk}.
Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \citep{Lequieu:2011fj} or even patient-to-patient heterogeneity \citep{2010_luan_varner_MolBioSys}.
Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \citep{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \citep{2012_creixell_linding_NatBiotech}.

@@

\subsection*{The model population captured key features of TGF-$\beta$ induced EMT}
The EMT model architecture, based upon curated molecular connectivity,
described the expression of 23 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
The EMT model contained 74 molecular species interconnected by 169 interactions.
Model equations were formulated as ordinary differential equations (ODEs) augmented with rule-based descriptions of activity and gene expression regulation.
ODEs are common tools to model biochemical pathways \citep{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
However, while ODE models can simulate complex intracellular behavior,
they require estimates for model parameters which are often difficult to obtain.
The EMT model had 251 unknown model parameters, 169 kinetic constants 38 control constants and 44 saturation constants.
As expected, these parameters were not uniquely identifiable given the training data \citep{Gadkar:2005vn}.
Thus, instead of identifying a single best fit (but uncertain) model,
we estimated a sloppy population of models (each consistent with the training data) by simultaneously minimizing the difference between model simulations and 41 molecular
data sets using the Pareto Optimal Ensemble Technique (JuPOETs).
The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
exposure to TGF-$\beta$ isoforms \citep{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by JuPOETs.
Additionally, we used data generated in this study (Fig. \ref{fg:VEGFA-Data}), and 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \citep{Lian:2011bs}.
To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
prediction error for each objective. Thus, we generated 11 different model ensembles.
Lastly, we compared model predictions with independent data sets not used during training (both at the molecular and model population
levels) to evaluate the predictive power of the parameter ensemble.

JuPOETs generated a population of probable signaling models which captured the multiple phases of EMT induction (Fig. \ref{fg:F2}).
JuPOETs sampled well over 10$^4$ probable models during each stage of the cross-validation using global random sampling.
From this analysis, N $\simeq$ 1400 models were selected for further analysis.
The selected models all had the same possible molecular connectivity, but different values for model parameters.
Transcription and translation rates, as well as mRNA and protein degradation terms, were set using physical values from the literature \citep{Milo:2010aa},
and allowed to vary by a scaling factor, see methods.
Model selection was based upon Pareto rank, the prediction and training error across all objectives.
The model population recapitulated key signaling events following TGF-$\beta$ exposure.
We subdivided the response to TGF-$\beta$ exposure into two phases.
First, TGF-$\beta$1/2 signaling initiated a program which downregulated E-cadeherin expression in a MAPK dependent manner while
simultaneously upregulating TGF-$\beta$3 expression. Second, TGF$-\beta$3 secretion initiated an autocrine feedback which upregulated the expression of
mesenchymal markers such as Vimentin and key upstream transcription factors such as LEF-1 in a SMAD dependent manner.
TGF-$\beta$3 expression was also able to sustain $\beta$-catenin release by inhibiting its sequestration by the APC complex through PI3K mediated GSK3,
which was captured by the model (Fig. \ref{fg:F4}B).
Each phase involved the hierarchal expression and/or post-translational modification of several key transcription factors.
During the first phase, stimulation with TGF-$\beta$1/2 (10 a.u.) activated both the SMAD and MAPK pathways.
MAPK activation resulted in the phosphorylation of the transcription factor activator protein 1 (AP-1), which in-turn upregulated the expression of Snail, a well established transcriptional repressor (Fig. \ref{fg:F2}A).
Snail expression was MAPK-dependent; the MEK inhibitor U0126 blocked AP-1 activation and Snail expression following TGF-$\beta$1/2 exposure (Fig. \ref{fg:F2}A, Lane 3).
Similar results were obtained for Slug expression, confirming initial activation through the MAPK pathway (data not shown).
Overexpression of either Snail or Slug upregulated TGF-$\beta$3 expression (Fig. \ref{fg:F2}C) while simultaneously downregulating E-cadeherin expression (Fig. \ref{fg:F2}F).
During the second phase, TGF$-\beta$3 secretion and the subsequent autocrine signaling resulted in the upregulation of mesenchymal marker expression.
The TGF$-\beta$3 induced gene expression program involves a complex hierarchy of transcriptional and post-translational regulatory events.
Absence of E-cadherin indirectly promoted TGF$-\beta$3 expression through the $\beta$-catenin/TCF4 complex following Snail or Slug expression (Fig. \ref{fg:F2}C, Lane 2 or 3).
Conversely, over-expression of E-cadherin inhibited the TGF$-\beta$3 autocrine production by sequestering cytosolic $\beta$-catenin, thereby blocking EMT (Fig. \ref{fg:F2}C, Lane 4 or 5).
TGF$-\beta$3 signaled through the Smad pathway to regulate LEF-1 expression and downstream target EMT genes (Fig. \ref{fg:F2}G).
TGF$-\beta$3 (10 a.u.) in combination with downstream inhibitors (DN-Smad4 and DN-LEF-1) completely inhibited Vimentin expression, while elevating E-cadherin expression (Fig. \ref{fg:F2}H,I).

The predictive power of the ensemble was tested using cross validation and by comparing simulations with data not used for model training.
In whole, all of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from a random starting point. Conversely, we \emph{predicted} all of the training objectives, at a 95\% confidence interval compared to randomized parameters (Wicoxon non-parametric test). The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 (not used for model training) to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L). Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \citep{Medici:2008fk} that drive the phenotypic conversion.
A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig. \ref{fg:ObjTable}).

>>>>

\documentclass[11pt,letterpaper]{article}
\usepackage[top=0.85in,left=2.75in,footskip=0.75in]{geometry}

% Use adjustwidth environment to exceed column width (see example table in text)
\usepackage{changepage}

% Use Unicode characters when possible
\usepackage[utf8]{inputenc}

% textcomp package and marvosym package for additional characters
\usepackage{textcomp,marvosym}

% fixltx2e package for \textsubscript
\usepackage{fixltx2e}

% amsmath and amssymb packages, useful for mathematical formulas and symbols
\usepackage{amsmath,amssymb}

% cite package, to clean up citations in the main text. Do not remove.
\usepackage{cite}

% Use nameref to cite supporting information files (see Supporting Information section for more info)
\usepackage{nameref,hyperref}

% line numbers
\usepackage[right]{lineno}

% ligatures disabled
\usepackage{microtype}
\DisableLigatures[f]{encoding = *, family = * }

% rotating package for sideways tables
\usepackage{rotating}

% Remove comment for double spacing
%\usepackage{setspace}
%\doublespacing

% Text layout
\raggedright
\setlength{\parindent}{0.5cm}
\textwidth 5.25in
\textheight 8.75in

% Bold the 'Figure #' in the caption and separate it from the title/caption with a period
% Captions will be left justified
\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}

% Use the PLoS provided BiBTeX style
\bibliographystyle{plos2009}

% Remove brackets from numbering in List of References
\makeatletter
\renewcommand{\@biblabel}[1]{\quad#1.}
\makeatother

% Leave date blank
\date{}

% Header and Footer with logo
\usepackage{lastpage,fancyhdr,graphicx}
\pagestyle{myheadings}
\pagestyle{fancy}
\fancyhf{}
\lhead{\includegraphics[natwidth=1.3in,natheight=0.4in]{PLOSlogo.png}}
\rfoot{\thepage/\pageref{LastPage}}
\renewcommand{\footrule}{\hrule height 2pt \vspace{2mm}}
\fancyheadoffset[L]{2.25in}
\fancyfootoffset[L]{2.25in}
\lfoot{\sf PLOS}

%% Include all macros below

\newcommand{\lorem}{{\bf LOREM}}
\newcommand{\ipsum}{{\bf IPSUM}}

\usepackage{listings}
\usepackage{colortbl}
\lstdefinestyle{myCustomMatlabStyle}{
%belowcaptionskip=\baselineskip,
breaklines=true,
%frame=L,
xleftmargin=\parindent,
language=Octave,
showstringspaces=false,
basicstyle=\tiny\ttfamily,
keywordstyle=\bfseries,
commentstyle=\itshape,
identifierstyle=\color{blue},
stringstyle=\color{orange},
}

%% END MACROS SECTION

\begin{document}
\vspace*{0.35in}
\begin{flushleft}
{\Large
\textbf\newline{Population heterogeneity in the epithelial to mesenchymal transition is controlled by NFAT and phosphorylated Sp1}}
\newline
% Insert Author names, affiliations and corresponding author email.
\\
Russell Gould\textsuperscript{1},
David Bassen\textsuperscript{1},
Anirikh Chakrabarti\textsuperscript{2},
Jeffrey D. Varner\textsuperscript{2},
Jonathan Butcher\textsuperscript{1,*},
\\
\bf{1} Nancy E. and Peter C. Meinig School of Biomedical Engineering, Cornell University, Ithaca NY
\\
\bf{2} Robert Frederick Smith School of Chemical and Biomolecular Engineering, Cornell University, Ithaca, NY, USA

% Insert additional author notes using the symbols described below. Insert symbol callouts after author names as necessary.
%
% Remove or comment out the author notes below if they aren't used.
%
% Primary Equal Contribution Note
%\Yinyang These authors contributed equally to this work.

% Additional Equal Contribution Note
%\ddag These authors also contributed equally to this work.

% Current address notes
%\textcurrency a Insert current address of first author with an address update
% \textcurrency b Insert current address of second author with an address update
% \textcurrency c Insert current address of third author with an address update

% Deceased author note
%\dag Deceased

% Group/Consortium Author Note
%\textpilcrow Insert Collaborative Author line here

* E-mail: jtb47@cornell.edu
\end{flushleft}

\section*{Abstract}
Epithelial to mesenchymal transition (EMT) is an essential differentiation program during tissue morphogenesis and remodeling.
EMT is induced by soluble transforming growth factor $\beta$ (TGF-$\beta$) family members, and restricted by vascular endothelial growth factor family members.
While many downstream molecular regulators of EMT have been identified, these have been largely evaluated individually without considering potential crosstalk.
In this study, we created an ensemble of dynamic mathematical models describing TGF-$\beta$ induced EMT to better understand the operational hierarchy of this complex molecular program.
We used ordinary differential equations (ODEs) to describe the transcriptional and post-translational regulatory events driving EMT.
Model parameters were estimated from multiple data sets using multiobjective optimization, in combination with cross-validation.
TGF-$\beta$ exposure drove the model population toward a mesenchymal phenotype, while an epithelial phenotype was enhanced following vascular endothelial growth factor A (VEGF-A) exposure.
Simulations predicted that the transcription factors phosphorylated SP1 and NFAT were master regulators promoting or inhibiting EMT, respectively.
Surprisingly, simulations also predicted that a cellular population could exhibit phenotypic heterogeneity (characterized by a significant fraction of the population with both high epithelial and mesenchymal marker expression)
if treated simultaneously with TGF-$\beta$ and VEGF-A.  We tested this prediction experimentally in both MCF10A and DLD1 cells and found that upwards of 45\% of the cellular population acquired this hybrid state in the presence of both TGF-$\beta$ and VEGF-A.
We experimentally validated the predicted NFAT/Sp1 signaling axis for each phenotype response. Lastly, we found that cells in the hybrid state had significantly different functional behavior when compared to VEGF-A or TGF-$\beta$ treatment alone.
Together, these results establish a predictive mechanistic model of EMT susceptibility, and potentially reveal a novel signaling axis which regulates carcinoma progression through an EMT versus tubulogenesis response.

@@

@@

\section*{Introduction}
The epithelial to mesenchymal transition (EMT) is a broadly participating, evolutionarily conserved differentiation program essential for tissue morphogenesis, remodeling and pathological processes such as cancer \cite{Thiery:2003il}. During EMT polarized, tightly adhered epithelial cell monolayers are transformed into non-interacting motile mesenchymal cells that simultaneously degrade and synthesize extracellular matrix (ECM) components and invade into the underlying tissue space \cite{Stahl:2001pd}. EMT is the fundamental initiator of developmental processes such as embryonic gastrulation and valvulogenesis \cite{Eisenberg:1995ai} (also Kalluri J Clin Invest 2009, Thiery Cell 2009). Transforming growth factor $\beta$ (TGF-$\beta$) family members are important inducers of both developmental and pathological EMT \cite{Zavadil:2005fu,Xu:2009oq}. Decades of research has focused on identifying molecular regulators of EMT, but almost all on a single gene and in a nearly binary yes/no level of qualitative understanding. Medici and coworkers identified a core signaling program by which TGF-$\beta$ isoforms induce EMT across a variety of cell lines \cite{Medici:2006qa,Medici:2008fk}.
This program involves carefully orchestrated rounds of gene expression driven by the Smad and Snail families of transcription factors as well as
other key factors such as lymphoid enhancer-binding factor 1 (LEF-1), nuclear factor of activated T-cells, cytoplasmic 1 (NFATc1), and specificity protein 1 (Sp1).
Coregulators such as $\beta$-catenin, NF-$\kappa$B, and the ErbB family of receptor tyrosine kinases however also participate in EMT regulation, but the degree of each's influence is difficult to ascertain in isolation \cite{Kim:2002lh,Jiang:2007aa,Huber:2004aa,Hardy:2010aa}.
EMT also exhibits complex temporal dynamics that are often intractable in gain/loss of function studies.
Elucidating the master regulatory architecture controlling EMT therefore requires inclusion of these complex overlapping and non-binary behaviors.

Systems biology and mathematical modeling are essential tools for understanding complex developmental programs like EMT \cite{Ahmed:2007mi}.
Previous computational models of TGF-$\beta$ induced differentiation focused on single biological factors or EMT in single cells.
For example, Chung \emph{et al.}, constructed a model of TGF-$\beta$ receptor activation and Smad signaling using ordinary differential equations and mass-action kinetics.
Their model suggested that a reduction of functional TGF-$\beta$ receptors in cancer cells may lead to an attenuated Smad2 signal \cite{Chung:2009jl}.
Similarly, Vilar \emph{et al.} suggested that specific changes in receptor trafficking patterns could lead to phenotypes that favor tumorigenesis \cite{Vilar:2006gb}.
Although these models provided insight into the role of receptor dynamics, EMT induction involves many other components, including competing second messengers and interconnected transcriptional regulatory loops.
Integrating these additional scales of molecular signaling while maintaining the capacity for robust prediction requires a new and expanded computational and experimental strategy.
Data-driven systems approaches \cite{Cirit:2012kx} or logical model formulations \cite{Morris:2011ys} are emerging paradigms that constrain model complexity through the incorporation of training and validation data.
These are interesting techniques because the data informs model structure (which can be expanded as more data becomes available).
Alternatively, Bailey proposed more than a decade ago that a qualitative understanding of a complex biological system should not require complete definition of its structural and parametric content \cite{2001_bailey_NatBiotech}.
Shortly thereafter, Sethna and coworkers showed that complex model behavior is often controlled by only a few parameter combinations, a characteristic seemingly universal to multi-parameter models referred to as ``sloppiness'' \cite{Machta:2013by}.
Thus, reasonable model predictions are often possible with only limited parameter information.
Taking advantage of this property, we developed sloppy techniques for parameter identification using ensembles of deterministic models \cite{Song:2010fk}.
Furthermore, we proposed that the sloppy behavior of biological networks may also be seen as a source of cell-to-cell \cite{Lequieu:2011fj} or even patient-to-patient heterogeneity \cite{2010_luan_varner_MolBioSys}.
Bayesian parameter identification techniques have also been used to explore cell-to-cell heterogeneity  \cite{2011_kalita_brasier_JBC,2011_hasenauer_allgower_BMCBioinfo},
where a population of cells could be viewed as a dynamic ensemble of context-specific biochemical networks \cite{2012_creixell_linding_NatBiotech}.

@@

\subsection*{The model population captured key features of TGF-$\beta$ induced EMT}
The EMT model architecture, based upon curated molecular connectivity,
described the expression of 23 genes following exposure to TGF-$\beta$ isoforms and VEGF-A (Fig. \ref{fg:F1}).
The EMT model contained 74 molecular species interconnected by 169 interactions.
Model equations were formulated as ordinary differential equations (ODEs) augmented with rule-based descriptions of activity and gene expression regulation.
ODEs are common tools to model biochemical pathways \cite{Schoeberl:2002rv,Chen:2009uq,Tasseff:2011ve}.
However, while ODE models can simulate complex intracellular behavior,
they require estimates for model parameters which are often difficult to obtain.
The EMT model had 251 unknown model parameters, 169 kinetic constants 38 control constants and 44 saturation constants.
As expected, these parameters were not uniquely identifiable given the training data \cite{Gadkar:2005vn}.
Thus, instead of identifying a single best fit (but uncertain) model,
we estimated a sloppy population of models (each consistent with the training data) by simultaneously minimizing the difference between model simulations and 41 molecular
data sets using the Pareto Optimal Ensemble Technique (JuPOETs).
The training data were generated in DLD1 colon carcinoma, MDCKII, and A375 melanoma cells following
exposure to TGF-$\beta$ isoforms \cite{Medici:2008fk}. We organized these data sets into 11 objective functions which were simultaneously minimized by JuPOETs.
Additionally, we used data generated in this study (Fig. \ref{fg:VEGFA-Data}), and 12 molecular data sets generated in HK-2 cells following VEGF-A exposure to train VEGF-A responsive model processes \cite{Lian:2011bs}.
To guard against overfitting, we augmented the multiobjective optimization with leave-one-out cross validation to independently estimate both the training and
prediction error for each objective. Thus, we generated 11 different model ensembles.
Lastly, we compared model predictions with independent data sets not used during training (both at the molecular and model population
levels) to evaluate the predictive power of the parameter ensemble.

JuPOETs generated a population of probable signaling models which captured the multiple phases of EMT induction (Fig. \ref{fg:F2}).
JuPOETs sampled well over 10$^4$ probable models during each stage of the cross-validation using global random sampling.
From this analysis, N $\simeq$ 1400 models were selected for further analysis.
The selected models all had the same possible molecular connectivity, but different values for model parameters.
Transcription and translation rates, as well as mRNA and protein degradation terms, were set using physical values from the literature \cite{Milo:2010aa},
and allowed to vary by a scaling factor, see methods.
Model selection was based upon Pareto rank, the prediction and training error across all objectives.
The model population recapitulated key signaling events following TGF-$\beta$ exposure.
We subdivided the response to TGF-$\beta$ exposure into two phases.
First, TGF-$\beta$1/2 signaling initiated a program which downregulated E-cadeherin expression in a MAPK dependent manner while
simultaneously upregulating TGF-$\beta$3 expression. Second, TGF$-\beta$3 secretion initiated an autocrine feedback which upregulated the expression of
mesenchymal markers such as Vimentin and key upstream transcription factors such as LEF-1 in a SMAD dependent manner.
TGF-$\beta$3 expression was also able to sustain $\beta$-catenin release by inhibiting its sequestration by the APC complex through PI3K mediated GSK3,
which was captured by the model (Fig. \ref{fg:F4}B).
Each phase involved the hierarchal expression and/or post-translational modification of several key transcription factors.
During the first phase, stimulation with TGF-$\beta$1/2 (10 a.u.) activated both the SMAD and MAPK pathways.
MAPK activation resulted in the phosphorylation of the transcription factor activator protein 1 (AP-1), which in-turn upregulated the expression of Snail, a well established transcriptional repressor (Fig. \ref{fg:F2}A).
Snail expression was MAPK-dependent; the MEK inhibitor U0126 blocked AP-1 activation and Snail expression following TGF-$\beta$1/2 exposure (Fig. \ref{fg:F2}A, Lane 3).
Similar results were obtained for Slug expression, confirming initial activation through the MAPK pathway (data not shown).
Overexpression of either Snail or Slug upregulated TGF-$\beta$3 expression (Fig. \ref{fg:F2}C) while simultaneously downregulating E-cadeherin expression (Fig. \ref{fg:F2}F).
During the second phase, TGF$-\beta$3 secretion and the subsequent autocrine signaling resulted in the upregulation of mesenchymal marker expression.
The TGF$-\beta$3 induced gene expression program involves a complex hierarchy of transcriptional and post-translational regulatory events.
Absence of E-cadherin indirectly promoted TGF$-\beta$3 expression through the $\beta$-catenin/TCF4 complex following Snail or Slug expression (Fig. \ref{fg:F2}C, Lane 2 or 3).
Conversely, over-expression of E-cadherin inhibited the TGF$-\beta$3 autocrine production by sequestering cytosolic $\beta$-catenin, thereby blocking EMT (Fig. \ref{fg:F2}C, Lane 4 or 5).
TGF$-\beta$3 signaled through the Smad pathway to regulate LEF-1 expression and downstream target EMT genes (Fig. \ref{fg:F2}G).
TGF$-\beta$3 (10 a.u.) in combination with downstream inhibitors (DN-Smad4 and DN-LEF-1) completely inhibited Vimentin expression, while elevating E-cadherin expression (Fig. \ref{fg:F2}H,I).

The predictive power of the ensemble was tested using cross validation and by comparing simulations with data not used for model training.
In whole, all of our training objectives were statistically significant (at a 95\% confidence interval) compared to a randomized parameter family (N = 100)
generated from a random starting point. Conversely, we \emph{predicted} all of the training objectives, at a 95\% confidence interval compared to randomized parameters (Wicoxon non-parametric test). The model also captured the temporal gene expression responses of E-cadherin, pSmad2, and LEF-1 (not used for model training) to within one-standard deviation (up to the 48 hr time-point) (Fig. \ref{fg:F2}J-L). Taken together, the model captured the key signaling events revealed by Medici \emph{et al.} \cite{Medici:2008fk} that drive the phenotypic conversion.
A listing of objective function values resulting from training, cross validation and the random parameter control is given in the supplement (Fig. \ref{fg:ObjTable}).
